Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for Pitching Motions Characteristic to Vertical Axis Turbines

### 2016-06-08 paper.tex
Modifications after Goude comments

\subsection{Actuator Line Model (ALM)} Is a three-dimensional and unsteady aerodynamic
model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
study the flow around wind turbines. It is a combination of a solver of
Navier-Stokes equations with a so-called actuator line technique, in which
blades of the turbine are represented by a radial distribution of body forces
along lines. These forces are determined using a dynamic stall model commonly
based on empirical data. This work uses the library turbinesFoam developed by
Bachant\cite{bachant2015simulating}. This model is based on incompressible Navier-Stokes equation
\begin{align}
 & \nonumber \\ & \dfrac{\partial \rho \vec{V}}{\partial t} + \vec{V} \cdot \nabla \vec{V} = - \dfrac{1}{ \rho} \nabla p + \nu \nabla ^2 \vec{V} + \vec{f}	  \label{eqmom}  
\end{align} %\begin{align}
with $\vec{f}$ representing the body forces, which are the loads on the rotating blades for this case.

Lift and drag coefficients of a 2D profile are already known, and these are tabulated and combined with the blade element approach for determining 
body forces acting on blades. Relative flow velocity $\vec{V}_{rel} $ and angle of attack $\alpha$ are calculated for each blade using the geometric relation between the velocity of the blade $- \omega r$, with $\omega$ representing the angular (rotational) velocity of the rotor and $r$ is the radius of the blade element, and the velocity of the local incoming wind flow $\vec{V}_{in}$. Figure \ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the plane perpendicular to the turbine axis.

@@

Once the angle of attack, relative velocity are obtained, the blade element lift and drag forces per spanwise length unit are calculated as
\begin{align}
& \vec{f}_L = \frac{1}{2} \ \rho \ c \ C_L \left| \vec{V}_{rel}^2 \right|  \label{lift}  
\end{align} %\begin{align}
\begin{align}
& \vec{f}_D = \frac{1}{2} \ \rho \ c \ C_D \left| \vec{V}_{rel}^2 \right|  \label{drag}  
\end{align} %\begin{align}
where $C_L$ and $C_D$ are the lift and drag coefficients, respectively. Both are in function of the Reynolds number and the angle of attack. It should be emphasized that the direction of the lift and drag are respectively perpendicular and parallel to the relative velocity $\vec{V}_{rel}$. $c$ represents the chord length.

@@

\subsection{Dynamic Stall Modeling (DSM)} Is represented by the Leishman-Beddoes
model\cite{leishman1986generalised} with the modifications of Sheng et
al\cite{sheng2008modified} and Dyachuk\cite{dyachuk}.
It is capable to calculate the unsteady lift,
pitching moment and drag, giving a physical description of the aerodynamics. It
have been validated with experimental data in\cite{leishman1989semi}.
The model is separated into three subsystems: an attached flow model for unsteady linear
airloads, a separated flow model for non-linear airloads and a dynamic stall
model for the airloads induced by the leading edge vortex.

The unsteady attached flow solution considers the circulatory and impulsive loads, which, at the same time depend on the unsteady bound vortex. The circulatory normal force coefficient could be expressed as
\begin{align}
& C_{N_n}^C = C_{N_\alpha} \alpha_{E_n}	\label{CNcirc}
\end{align}
where $C_{N \alpha}$ indicates the slope of the static normal force coefficient for a specific Reynolds number, and $\alpha_{E_n}$ is an expression for an equivalent angle of attack
\begin{align}
& \alpha_{E_n} = \alpha_n - X_n -Y_n - Z_n	\label{alphaeq}
\end{align}
where $\alpha$ is the geometrical angle of attackm $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pithing rate, the details are found in \cite{dyachuk2013dynamic}. Indices $n$ and $n-1$ in equations \ref{CNcirc} and \ref{alphaeq} represent the current and previous time-step. A delayed angle of attack is considered due to the lag in pressure response, and it is calculated as
\begin{align}
& {\alpha _n}' = {\alpha _n} - D_{\alpha _n}    	\label{alphadel}
\end{align}
with $D_\alpha$ as the deficiency function
\begin{align}
& D_{\alpha_n} = D_{\alpha_{n-1}} \exp \left( - \frac{\Delta s}{T_\alpha} \right) + (\alpha_n - \alpha_{n-1})\exp \left( - \frac{\Delta s}{2T_\alpha} \right) \label{defalpha}
\end{align}
with the empirically derived time constant $T_\alpha$, which has a value of $T_\alpha$ = 6.3 for the NACA0021 airfoil, and $\Delta s$ corresponds to a non-dimensional time-step
\begin{align}
& \Delta s| = \frac{2 | \vec{V}_{rel} | \Delta t}{c}		.	\label{deltas}
\end{align}

Because of the reversal flow in the boundary layer, a leading edge vortex is created at the surface of the airfoil. The critical angle of attack $\alpha_{cr_n}$ defines the condition at which dynamic stall begins
\begin{align}
\alpha_{cr_n} =
\begin{cases}
    \alpha_{ds0}       & \quad |r_n| \geqslant r_0 \\
    \alpha_{ss} + (\alpha_{ds0} - \alpha_{ss}) \frac{| r_n |}{r_0}       & \quad |r_n| < r_0 \\
\end{cases}		\label{criticalpha}
\end{align}
with the reduced pitching rate $r_n$ defined as
\begin{align}
r_n = \frac{\dot{\alpha}_n c}{ 2 | \vec{V}_{rel} | }
\end{align}
here, $\dot{\alpha}$ is the pitch rate, $r_0$ is the reduced pitching rate, which limits the quasi-steady stall with the dynamic stall. Its value is $r_0$ = 0.01 for the NACA0021 airfoil. The static stall onset angle $\alpha_{ss}$ and the critical stall onset angle $\alpha_{ds0}$ are 14.33 and 17.91, respectively. The following is the the dynamic stall condition used
\begin{align}
| \alpha ' | > \alpha_{cr} \rightarrow \mathrm{stall} \label{stall}
\end{align}
for the separated flow, the effects are divided in two groups: trailing edge separation and leading vortex convection. The first one is related to the time delay in the movement of the boundary separation point, and is calculated using Kirchhoff's approximation  
\begin{align}
f'_n =
\begin{cases}
    1- 0.4 \exp \left( \frac{|\alpha'_n| - \alpha_1 }{S_1}   \right)    & \quad |\alpha'_n| <  \alpha_1 \\
    0.02- 0.58 \exp \left( \frac{\alpha_1 - |\alpha'_n| }{S_2} \right)  & \quad |\alpha'_n| \geqslant  \alpha_1 \\
\end{cases}		\label{criticalpha2}
\end{align}
where $f'$ is the delayed separation point and $\alpha_1$, $S_1$ and $S_2$ are constant based on the airfoil profile and the local Reynolds number, these can be found in \cite{dyachuk2013dynamic}. The boundary layer around the blade is in function on the time, and this effect is superimposed on the pressure response delay. The additional delay is represented by $f''$, which is the dynamic separation point
\begin{align}
f''_n = f'_n - D _{f_n}	   \label{dyndelay}
\end{align}
and $D_{f_n}$ is the deficiency function defined as
\begin{align}
& D_{f_n} = D_{f_{n-1}} \exp \left( - \frac{\Delta s}{T_f} \right) + (f_n - f_{n-1})\exp \left( - \frac{\Delta s}{2T_f} \right) \label{deff}
\end{align}
here, $T_f$ is empirically derived time constant and its value is $T_f$ = 3. Then, the normal force coefficient for unsteady conditions before the dynamic stall onset is calculated as
\begin{align}
& C_{N_n}^f = C_{N_\alpha} \alpha_{E_n} \left( \frac{1 + \sqrt{f''}}{2} \right)^2	\label{CNf}
\end{align}
after the stall condition is met, the leading edge vortex convects over the surface of the airfoil towards the trailing edge. A significant increase in the normal forces occurs during this process
\begin{align}
C_N^v = B_1 (f'' - f) V_x \label{CNv}
\end{align}
where $C_N^v$ represents the normal force coefficient during the vortex convection, which is in function of the pitch rate. $V_x$ and $B_1$ are parameters based on the local Reynolds number and the airfoil profile, they are found in \cite{dyachuk2013dynamic}. Once the vortex passes the trailing edge, the normal force decreases quickly. Total normal force obtained is expressed as
\begin{align}
C_N = C_{N_n}^f + C_N^v \label{CNtotal}
\end{align}
Figure \ref{figpitching} shows an example of the normal coefficient forces on a blade during pitching motion, using the above model described.

@@

The model is more sensitive to the spatial discretization than the temporal one; same characteristic was obtained by Bachant et al. in \cite{Bachant2016-VAT-ALM}. Accuracy and stability were not affected even for big values for $\Delta t$, as long they are within the limit given by the CFL condition; a Courant number of less than unity is required. On the other hand, it should be considered that small time-step of discretization could cause numerical instabilities due to the fluctuation of the flow fields resolving the transient term $\frac{\partial}{\partial t}$.

@@

For a TSR of 3.34 (Figure \ref{fig174}), the maximum magnitude of the angle of attack is at
17.4\degree\ which is in the stall region. Both \textit{CN} and \textit{CT} curves show the delay
of the flow reattachment, which is characteristic of the dynamic stall
phenomenon. Peaks of simulated values have close agreement with the measured
data. The stall onset angle flow reattachment are well predicted.

@@

\subsection{Turbulence model comparison}
A simulation with a RANS $k$-$\epsilon$ standard turbulence model was carried out to compare the obtained results against another one using a LES Smagorinsky turbulence model. Figure \ref{RANSLES} shows that for the $C_N$ values there is not considerable difference, but a small improvement has been achieved for the $C_T$ values. Another study cases conducted for a whole turbine simulation showed that the LES turbulence model gives better results than RANS.

>>>>

\subsection{Actuator Line Model (ALM)} 
The ALM is a three-dimensional and unsteady aerodynamic
model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
study the flow around wind turbines. It is a combination of a solver of
Navier-Stokes equations with a so-called actuator line technique, in which
blades of the turbine are represented by a radial distribution of body forces
along lines. These forces are determined using a dynamic stall model commonly
based on empirical data. This work uses the library turbinesFoam developed by
Bachant\cite{bachant2015simulating}. This model is based on incompressible Navier-Stokes equation
\begin{align}
 & \nonumber \\ & \dfrac{\partial \vec{V}}{\partial t} + \vec{V} \cdot \nabla \vec{V} = - \dfrac{1}{ \rho} \nabla p + \nu \nabla ^2 \vec{V} + \vec{f}	  \label{eqmom}  
\end{align} %\begin{align}
with $\vec{f}$ representing the body forces, which are the loads on the rotating blades for this case.

Static lift and drag coefficients of the 2D profile used for this study are already known from the technical technical report made by Sheldahl and Klimas \cite{sheldahl1981aerodynamic}, which is a popular database set containing values for a wide range of Reynolds numbers. These coefficients are tabulated and combined with the blade element approach for determining 
body forces acting on blades. Relative flow velocity $\vec{V}_{rel} $ and angle of attack $\alpha$ are calculated for each blade using the geometric relation between the velocity of the blade $- \Omega r$, with $\omega$ representing the angular (rotational) velocity of the rotor and $r$ is the radius of the blade element, and the velocity of the local incoming wind flow $\vec{V}_{in}$. Figure \ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the plane perpendicular to the turbine axis.

@@

Once the angle of attack, relative velocity are obtained, the blade element lift and drag forces per spanwise length unit are calculated as
\begin{align}
& f_L = \frac{1}{2} \ \rho \ c \ C_L \left| V_{rel} \right|^2  \label{lift}  
\end{align} %\begin{align}
\begin{align}
& f_D = \frac{1}{2} \ \rho \ c \ C_D \left| V_{rel} \right|^2  \label{drag}  
\end{align} %\begin{align}
where $C_L$ and $C_D$ are the lift and drag coefficients, respectively. Both are function of the Reynolds number and the angle of attack. It should be emphasized that the direction of the lift and drag are respectively perpendicular and parallel to the relative velocity $\vec{V}_{rel}$. $c$ represents the chord length.

@@

\subsection{Dynamic Stall Modeling (DSM)} The DSM is represented by the Leishman-Beddoes
model\cite{leishman1986generalised} with the modifications of Sheng et
al\cite{sheng2008modified} and Dyachuk\cite{dyachuk}.
It is capable to calculate the unsteady lift,
pitching moment and drag, giving a physical description of the aerodynamics. It
have been validated with experimental data in\cite{leishman1989semi}.
The model is separated into three subsystems: an attached flow model for unsteady linear
airloads, a separated flow model for non-linear airloads and a dynamic stall
model for the airloads induced by the leading edge vortex.

The unsteady attached flow solution considers the circulatory and impulsive loads, which, at the same time depend on the unsteady bound vortex. The circulatory normal force coefficient could be expressed as
\begin{align}
& C_{N_n}^C = C_{N_\alpha} \alpha_{E_n}	\label{CNcirc}
\end{align}
where $C_{N \alpha}$ indicates the slope of the static normal force coefficient for a specific Reynolds number, and $\alpha_{E_n}$ is an expression for an equivalent angle of attack
\begin{align}
& \alpha_{E_n} = \alpha_n - X_n -Y_n - Z_n	\label{alphaeq}
\end{align}
where $\alpha$ is the geometrical angle of attackm $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pithing rate, the details are found in \cite{dyachuk2013dynamic}. Indices $n$ and $n-1$ in equations \ref{CNcirc} and \ref{alphaeq} represent the current and previous time-step. A delayed angle of attack is considered due to the lag in pressure response, and it is calculated as
\begin{align}
& {\alpha _n}' = {\alpha _n} - D_{\alpha _n}    	\label{alphadel}
\end{align}
with $D_\alpha$ as the deficiency function
\begin{align}
& D_{\alpha_n} = D_{\alpha_{n-1}} \exp \left( - \frac{\Delta s}{T_\alpha} \right) + (\alpha_n - \alpha_{n-1})\exp \left( - \frac{\Delta s}{2T_\alpha} \right) \label{defalpha}
\end{align}
with the empirically derived time constant $T_\alpha$, which has a value of $T_\alpha$ = 6.3 for the NACA0021 airfoil, and $\Delta s$ corresponds to a non-dimensional time-step
\begin{align}
& \Delta s = \frac{2 | \vec{V}_{rel} | \Delta t}{c}		.	\label{deltas}
\end{align}

Because of the reversal flow in the boundary layer, a leading edge vortex is created at the surface of the airfoil. The critical angle of attack $\alpha_{cr_n}$ defines the condition at which dynamic stall begins
\begin{align}
\alpha_{cr_n} =
\begin{cases}
    \alpha_{ds0}       & \quad |r_n| \geqslant r_0 \\
    \alpha_{ss} + (\alpha_{ds0} - \alpha_{ss}) \frac{| r_n |}{r_0}       & \quad |r_n| < r_0 \\
\end{cases}		\label{criticalpha}
\end{align}
with the reduced pitching rate $r_n$ defined as
\begin{align}
r_n = \frac{\dot{\alpha}_n c}{ 2 | \vec{V}_{rel} | }
\end{align}
here, $\dot{\alpha}$ is the pitch rate, $r_0$ is the reduced pitching rate, which limits the quasi-steady stall with the dynamic stall. Its value is $r_0$ = 0.01 for the NACA0021 airfoil. The static stall onset angle $\alpha_{ss}$ and the critical stall onset angle $\alpha_{ds0}$ are 14.33 and 17.91, respectively. The following is the the dynamic stall condition used
\begin{align}
| \alpha ' | > \alpha_{cr} \rightarrow \mathrm{stall} \label{stall}
\end{align}
for the separated flow, the effects are divided in two groups: trailing edge separation and leading vortex convection. The first one is related to the time delay in the movement of the boundary separation point, and is calculated using Kirchhoff's approximation  
\begin{align}
f'_n =
\begin{cases}
    1- 0.4 \exp \left( \frac{|\alpha'_n| - \alpha_1 }{S_1}   \right)    & \quad |\alpha'_n| <  \alpha_1 \\
    0.02- 0.58 \exp \left( \frac{\alpha_1 - |\alpha'_n| }{S_2} \right)  & \quad |\alpha'_n| \geqslant  \alpha_1 \\
\end{cases}		\label{criticalpha2}
\end{align}
where $f'$ is the delayed separation point and $\alpha_1$, $S_1$ and $S_2$ are constant based on the airfoil profile and the local Reynolds number, these can be found in \cite{dyachuk2013dynamic}. The boundary layer around the blade is in function on the time, and this effect is superimposed on the pressure response delay. The additional delay is represented by $f''$, which is the dynamic separation point
\begin{align}
f''_n = f'_n - D _{f_n}	   \label{dyndelay}
\end{align}
and $D_{f_n}$ is the deficiency function defined as
\begin{align}
& D_{f_n} = D_{f_{n-1}} \exp \left( - \frac{\Delta s}{T_f} \right) + (f_n - f_{n-1})\exp \left( - \frac{\Delta s}{2T_f} \right) \label{deff}
\end{align}
Here, $T_f$ is empirically derived time constant and its value is $T_f$ = 3. Then, the normal force coefficient for unsteady conditions before the dynamic stall onset is calculated as
\begin{align}
& C_{N_n}^f = C_{N_\alpha} \alpha_{E_n} \left( \frac{1 + \sqrt{f''}}{2} \right)^2	\label{CNf}
\end{align}
after the stall condition is met, the leading edge vortex convects over the surface of the airfoil towards the trailing edge. A significant increase in the normal forces occurs during this process
\begin{align}
C_N^v = B_1 (f'' - f) V_x \label{CNv}
\end{align}
where $C_N^v$ represents the normal force coefficient during the vortex convection, which is in function of the pitch rate. $V_x$ and $B_1$ are parameters based on the local Reynolds number and the airfoil profile, they are found in \cite{dyachuk2013dynamic}. Once the vortex passes the trailing edge, the normal force decreases quickly. Total normal force obtained is expressed as
\begin{align}
C_N = C_{N_n}^f + C_N^v \label{CNtotal}
\end{align}
Figure \ref{figpitching} shows an example of the normal coefficient forces on a blade during pitching motion, using the above model described.

@@

The model is more sensitive to the spatial discretization than the temporal one; same characteristic was obtained by Bachant et al. in \cite{Bachant2016-VAT-ALM}. Accuracy and stability were not affected even for big values for $\Delta t$, as long they are within the limit given by the CFL condition; a Courant number of less than unity is required. On the other hand, it should be considered that small time-step of discretization could cause numerical instabilities due to the fluctuation of the flow fields resolving the transient term $\frac{\partial \alpha}{\partial t}$.

@@

For a TSR of 3.34 (Figure \ref{fig174}), the maximum magnitude of the angle of attack is at
17.4\degree\ which is in the stall region. Both $C_N$ and $C_T$ curves show the delay
of the flow reattachment, which is characteristic of the dynamic stall
phenomenon. Peaks of simulated values have close agreement with the measured
data. The stall onset angle flow reattachment are well predicted.

@@

\subsection{Turbulence model comparison}
A simulation with a RANS $k$-$\epsilon$ standard turbulence model was carried out to compare the obtained results against another one using a LES Smagorinsky turbulence model. Figure \ref{RANSLES} shows that for the $C_N$ values there is not considerable difference, but a small improvement has been achieved for the $C_T$ values.
